Fast Regression of the Tritium Breeding Ratio in Tokamak Fusion Reactors

### 2020-05-29 paper/conclusion.tex
more edits, add conclusion and acknowledgement, reformat QASS figures
Reduce introduction and adaptive sampling sections.

Over the course of this project, we employed a broad spectrum of data
analysis and machine learning techniques to develop fast and high-quality
surrogates for a~MC TBR simulation in use at~UKAEA. Having implemented a sampling
software to efficiently evaluate this expensive MC~model, we deployed it on a
high performance cluster to generate over~\num{900000}
datapoints for training and test purposes. We
investigated possibilities for simplification of the parameter space, and
concluded that no straightforward reduction was possible. After reviewing
9~surrogate model families, examining their behaviour on constrained and
unrestricted feature space, and studying their scaling properties, we retrained
the best-performing instances to produce properties desirable for
practical use. The fastest surrogate, an artificial neural network trained
on~\num{500000} datapoints, featured~$R^2=\num{0.985}$ with mean prediction time
of~$\SI{0.898}{\micro\second}$, representing a relative
speedup of $8\cdot 10^6$ with respect to the MC model. Alternatively, we
also demonstrated the possibility of achieving comparable results using only a
training set of size~\num{10000}.

After a thorough review of the literature, we developed a novel adaptive
sampling algorithm, QASS, capable of interfacing with any of the studied
surrogates. Preliminary testing on a toy theory, qualitatively comparable to
the MC TBR model, demonstrated the effectiveness of QASS and behavioural trends
consistent with the design of the algorithm. With~\num{100000} initial samples and 100 incremental samples per iteration, QASS achieved a ${\sim}40\%$ decrease in surrogate error as compared with a baseline random sampling scheme. Equivalently, the same surrogate error as the baseline could be achieved with ${\sim}6\%$ fewer total samples of the expensive theory. Further optimisation over the hyperparameter space has strong
potential to increase this performance, in particular by decreasing the required quantity of initial samples. This will allow for future deployment of QASS
in coalition with any of the most effective identified surrogates to facilitate
learning during evaluation of the MC TBR model.

\section*{Acknowledgements}

The authors would like to thank Simeon Bash, Vignesh Gopakumar, Prof.~Nikolaos
Konstantinidis, Nikolaos Nikolaou, Prof.~Emily Nurse, Jonathan Shimwell and Ingo
Waldmann for their supervision and valuable suggestions related to this work.

>>>>

We employed a broad spectrum of data
analysis and machine learning techniques to develop fast and high-quality
surrogates for the Paramak TBR model. After reviewing
nine surrogate model families, examining their behaviour on constrained and
unrestricted feature space, and studying their scaling properties, we retrained
the best-performing instances to produce properties desirable for
practical use. The fastest surrogate, an artificial neural network trained
on~\num{500000} datapoints, featured~$R^2=\num{0.985}$ with mean prediction time
of~$\SI{0.898}{\micro\second}$, representing a relative
speedup of $8\cdot 10^6$ with respect to the MC model. We demonstrated the possibility of achieving comparable results using only a
training set of size~\num{10000}.

We further developed a novel adaptive
sampling algorithm, QASS, capable of interfacing with any of the studied
surrogates. Preliminary testing on a toy theory, qualitatively comparable to
the MC TBR model, demonstrated the effectiveness of QASS and behavioural trends
consistent with the design of the algorithm. With~\num{100000} initial samples and 100 incremental samples per iteration, QASS achieved a ${\sim}40\%$ decrease in surrogate error as compared with a baseline random sampling scheme. Further optimisation over the hyperparameter space has strong potential to increase this performance, in particular by decreasing the required quantity of initial samples. This will allow for future deployment of QASS on the Paramak TBR model in coalition with any of the most effective identified surrogates.

### 2020-05-29 paper/introduction.tex
Reduce introduction and adaptive sampling sections.

The analysis of massive datasets has become a necessary component of virtually all technical fields, as well as the social and humanistic sciences, in recent years. Given that rapid improvements in sensing and processing hardware have gone hand in hand with the data explosion, it is unsurprising that software for the generation and interpretation of this data has also attained a new frontier in complexity. In particular, simulation procedures such as Monte Carlo (MC) event generation can perform physics predictions even for theoretical regimes which are not analytically tractable. The bottleneck for such procedures, as is often the case, lies in the computational time and power which they necessitate.

Surrogate models, or metamodels, can resolve this limitation by replacing a
resource-expensive procedure with a much cheaper approximation
\cite{Sondergaard2003}. They are especially useful in applications where
numerous evaluations of an expensive procedure are required over the same or
similar domains, e.g. in the parameter optimisation of a theoretical model. The
term ``metamodel'' proves especially meaningful in this case, when the surrogate
model approximates a computational process which is itself a model for a
(perhaps unknown) physical process~\cite{Myers2002}. There exists a spectrum
between ``physical'' surrogates which are constructed with some contextual
knowledge in hand, and ``empirical'' surrogates which are derived purely from
samples of the underlying expensive model.

In this project, in coordination with the UK Atomic Energy Authority (UKAEA), we sought to develop a surrogate model for the tritium breeding ratio (TBR) in a Tokamak nuclear fusion reactor. Our expensive model was an MC-based neutronics simulation, \textit{Paramak}\footnote{Provided by collaborator Jonathan Shimwell, at UKAEA.}, which returns a prediction of the TBR for a given configuration of a spherical Tokamak. We took an empirical approach to the construction of this surrogate, and no results described here are explicitly dependent on prior physics knowledge.

For the remainder of Section 1, we will define the TBR and set the context of this work within the goals of the UKAEA. In Section 2 we will describe our datasets generated from the expensive model for training and validation purposes, and the dimensionality reduction methods employed to develop our understanding of the parameter domain. In Section 3 we will present our methodologies for the comparison testing of a wide variety of surrogate modelling techniques, as well as a novel adaptive sampling procedure suited to this application. After delivering the results of these approaches in Section 4, we will give our final conclusions and recommendations for further work.

@@

Nuclear fusion technology relies on the production and containment of an
extremely hot and dense plasma. In this environment, by design similar to that
of a star, hydrogen atoms attain energies sufficient to overcome their usual
electrostatic repulsion and fuse to form helium~\cite{Hernandez2018}. Early prototype reactors
made use of the deuterium (\isotope[2]{H}, or \isotope{D}) isotope of hydrogen in order to
achieve fusion under more accessible conditions, but lead to limited success.
The current frontier generation of fusion reactors, such as the Joint European Torus (JET) and the
under-construction International Thermonuclear Experimental Reactor (ITER), make
use of tritium (\isotope[3]{H}, or \isotope{T}) fuel for further efficiency gain.
Experimentation at JET dating back to 1997~\cite{Keilhacker1999} has made significant headway in
validating deuterium-tritium (D-T) operations and constraining the technology
which will be employed in ITER in a scaled-up form.

However, tritium is much less readily available as a fuel source than deuterium.
While at least one deuterium atom occurs for every \num{5000} molecules of
naturally-sourced water, and may be easily distilled, tritium is extremely rare
in nature. It may be produced indirectly through irradiation of heavy water
(\DDO) during nuclear fission, but only at very low rates which could
never sustain industrial-scale fusion power.

Instead, modern D-T reactors rely on tritium breeding blankets, specialised
layers of material which partially line the reactor and produce tritium upon
neutron bombardment, e.g. by 
\begin{eqnarray}
	\isotope[1][0]{n} + \hspace{3pt} \isotope[6][3]{Li} 
	&\longrightarrow \hspace{3pt} 
	\isotope[3][1]{T} + \hspace{3pt}\isotope[4][2]{He} \\
	\isotope[1][0]{n} + \hspace{3pt} \isotope[7][3]{Li} 
	&\longrightarrow \hspace{3pt} 
	\isotope[3][1]{T} + \hspace{3pt} \isotope[4][2]{He} + \hspace{3pt} \isotope[1][0]{n}
\end{eqnarray}
\begin{figure}
  \centering
  % TODO: \includegraphics[width=0.48\textwidth]{fig1_tokamakdiagram.png}
  \caption{Typical single-null reactor configuration as specified by BLUEPRINT~\cite{Coleman2019}: 1 — plasma,
2 — breeding blankets }
\label{fig:tokamak}
\end{figure}
where T represents tritium and \isotope[7]{Li}, \isotope[6]{Li} are the more and
less frequently occurring isotopes of lithium, respectively. \isotope[6]{Li} has
the greatest tritium breeding cross-section of all tested isotopes~\cite{Hernandez2018}, but due
to magnetohydrodynamic instability of liquid lithium in the reactor environment,
a variety of solid lithium compounds are preferred.

The TBR is defined as the ratio between tritium generation in the breeding
blanket per unit time and tritium fuel consumption in the reactor. The MC
neutronics simulations previously mentioned therefore must account for both the
internal plasma dynamics of the fusion reactor and the resultant interactions of
neutrons with breeding blanket materials. Neutron paths are traced through a CAD
model (e.g. Figure~\ref{fig:tokamak}) of a reactor with modifiable geometry.

The input parameters of the computationally-expensive TBR model therefore fall
into two classes. Continuous parameters, including material thicknesses and
packing ratios, describe the geometry of a given reactor configuration. Discrete
categorical parameters further specify all relevant material sections, including
coolants, armours, and neutron multipliers. One notable exception is the
enrichment ratio, a continuous parameter denoting the presence
of~\isotope[6]{Li}. Our challenge, put simply, was to produce a fast TBR
function which takes these same input parameters and approximates the MC TBR
model with the greatest achievable regression performance.

>>>>

Surrogate models were developed to resolve computational limitations in the analysis of massive datasets by replacing a resource-expensive procedure with a much cheaper approximation
\cite{Sondergaard2003}. They are especially useful in applications where
numerous evaluations of an expensive procedure are required over the same or
similar domains, e.g. in the parameter optimisation of a theoretical model. The
term ``metamodel'' proves especially meaningful in this case, when the surrogate
model approximates a computational process which is itself a model for a
(perhaps unknown) physical process~\cite{Myers2002}. There exists a spectrum
between ``physical'' surrogates which are constructed with some contextual
knowledge in hand, and ``empirical'' surrogates which are derived purely from
samples of the underlying expensive model.

We sought to develop an empirical surrogate model for the tritium breeding ratio (TBR) in a Tokamak nuclear fusion reactor. Our expensive model was a Monte Carlo (MC) neutronics simulation, \textit{Paramak}\footnote{Provided by collaborator Jonathan Shimwell, at UKAEA.}, which returns a prediction of the TBR for a given configuration of a spherical Tokamak.

For the remainder of Section 1, we will define the TBR and further motivate our research. In Section 2 we will present our methodologies for the comparison testing of a wide variety of surrogate modelling techniques, as well as a novel adaptive sampling procedure suited to this application. After delivering the results of these approaches in Section 3, we will give our final conclusions and recommendations for further work.

@@

Nuclear fusion technology relies on the production and containment of an
extremely hot and dense plasma containing enriched Hydrogen isotopes. The current frontier generation of fusion reactors, such as the Joint European Torus (JET) and the
under-construction International Thermonuclear Experimental Reactor (ITER), make
use of both tritium and deuterium fuel. While at least one deuterium atom occurs for every \num{5000} molecules of naturally-sourced water, and may be easily distilled, tritium is extremely rare in nature. It may be produced indirectly through irradiation of heavy water
(\DDO) during nuclear fission, but only at very low rates which could
never sustain industrial-scale fusion power.

Instead, modern D-T reactors rely on tritium breeding blankets, specialised
layers of material which partially line the reactor and produce tritium upon
neutron bombardment, e.g. by 
\begin{eqnarray}
	\isotope[1][0]{n} + \hspace{3pt} \isotope[6][3]{Li} 
	&\longrightarrow \hspace{3pt} 
	\isotope[3][1]{T} + \hspace{3pt}\isotope[4][2]{He} \\
	\isotope[1][0]{n} + \hspace{3pt} \isotope[7][3]{Li} 
	&\longrightarrow \hspace{3pt} 
	\isotope[3][1]{T} + \hspace{3pt} \isotope[4][2]{He} + \hspace{3pt} \isotope[1][0]{n}
\end{eqnarray}

where T represents tritium and \isotope[7]{Li}, \isotope[6]{Li} are the more and
less common isotopes of lithium, respectively. The TBR is defined as the ratio between tritium generation in the breeding blanket and consumption in the reactor, whose input parameters in the-expensive MC model fall
into two classes. Continuous parameters, including material thicknesses and
packing ratios, describe the geometry of a given reactor configuration. Discrete
categorical parameters then specify material selections, including
coolants, armours, and neutron multipliers. We set out to produce a fast TBR
function which takes these same input parameters and approximates the MC TBR
model with the greatest achievable regression performance.

### 2020-05-29 paper/methodology.tex
Reduce introduction and adaptive sampling sections.

For both methods, we selected state-of-the-art regression algorithms to perform
surrogate training on sampled point sets. Listed in~\tref{tbl:surrogates}, these
implementations define 9~surrogate families that are later reviewed in~\sref{sec:results}.
We note that each presented algorithm defines hyperparameters that may influence its
performance. Since their optimal values for this problem are unknown, we explore
their assignments prior to other experiments.

@@

\subsection{Adaptive Approach}\label{sec:adaptive}
\todo{This section needs to be shortened}

\begin{figure}
	\centering
	\includegraphics[width=\linewidth]{fig4_qassplan.png}
	\caption{\label{fig:qassplan}Schematic of QASS algorithm}
\end{figure}

All of the surrogate modelling techniques studied in this project face a common
challenge: their accuracy is limited by the quantity of training samples which
are available from the expensive MC TBR model. Adaptive sampling procedures can
improve upon this limitation by taking advantage of statistical information
which is accumulated during the training of any surrogate model. Rather than
training the surrogate on a single sample set generated according to a fixed
strategy, sample locations are chosen periodically during training so as to best suit the model
under consideration.

Adaptive sampling techniques appear frequently in the literature and have been
specialised for surrogate modelling. Garud's~\cite{Garud2016} ``Smart Sampling
Algorithm'' achieved notable success by incorporating surrogate quality and
crowding distance scoring to identify optimal new samples, but was only tested
on a single-parameter domain. We theorised that a nondeterministic sample
generation approach, built around Markov Chain Monte Carlo methods (MCMC), would
fare better for high-dimensional models by more thoroughly exploring all local
optima in the feature space. MCMC produces a progressive chain of sample points,
each drawn according to the same symmetric proposal
distribution\todo{Incorporate the footnote into the text.}\footnote{An
adaptive MCMC procedure~\cite{Zhang2012}, which adjusts an ellipsoidal proposal
distribution to fit the posterior, was also implemented but not fully tested.}
from the prior point. These sample points will converge to a desired posterior
distribution, so long as the acceptance probability for these draws has a
particular functional dependence on that posterior value (see~\cite{Zhou2018}
for a review).

Many researchers have embedded surrogate methods into MCMC strategies for
parameter optimisation~\cite{Zhang2020,Gong2017}, in particular the ASMO-PODE
algorithm~\cite{Ginting2011} which makes use of MCMC-based adaptive sampling to
attain greater surrogate precision around prospective optima. Our novel approach
draws inspiration from ASMO-PODE, but instead uses MCMC to generate samples
which increase surrogate precision throughout the entire parameter space.

@@

% Adaptive multi-chain MCMC:~\cite{Zhang2012}

>>>>

For both methods, we selected state-of-the-art regression algorithms to perform
surrogate training on sampled point sets. Listed in~\tref{tbl:surrogates}, these
implementations define nine surrogate families that are later reviewed in~\sref{sec:results}.
We note that each presented algorithm defines hyperparameters that may influence its
performance. Since their optimal values for this problem are unknown, we explore
their assignments prior to other experiments.

@@

\subsection{Adaptive Approach}\label{sec:adaptive}

\begin{figure}
	\centering
	\includegraphics[width=1.2\linewidth]{fig4_qassplan.png}
	\caption{\label{fig:qassplan}Schematic of QASS algorithm}
\end{figure}

Adaptive sampling techniques appear frequently in the literature and have been specialised for surrogate modelling, where precision is implicitly limited by the quantity of training samples which are available from the expensive model. Garud's~\cite{Garud2016} ``Smart Sampling Algorithm'' achieved notable success by incorporating surrogate quality and
crowding distance scoring to identify optimal new samples, but was only tested
on a single-parameter domain. We theorised that a nondeterministic sample
generation approach, built around Markov Chain Monte Carlo methods (MCMC), would
fare better for high-dimensional models by more thoroughly exploring all local
optima in the feature space. MCMC produces each sample points according to a shared proposal
distribution from the prior point. These sample points will converge to a desired posterior
distribution, so long as the acceptance probability meets certain statistical criteria (see~\cite{Zhou2018} for a review).

Many researchers have embedded surrogate methods into MCMC strategies for
parameter optimisation~\cite{Zhang2020,Gong2017}, in particular the ASMO-PODE
algorithm~\cite{Ginting2011} which makes use of MCMC-based adaptive sampling. Our approach draws inspiration from ASMO-PODE, but instead uses MCMC to generate samples
which increase surrogate precision throughout the entire parameter space.

@@

### 2020-05-29 paper/results.tex
Reduce introduction and adaptive sampling sections.

\todo{Need any introduction here?}

@@

\subsection{Adaptive Approach}\label{sec:adaptiveres}
\todo{This section needs to be shortened.}

\begin{figure}
	\centering
	\includegraphics[width=0.9\linewidth]{fig5_sintoy}
	\caption{Sinusoidal toy TBR theory over 2 continuous parameters,
	$n=1$.}
	\label{fig:sintoy}
\end{figure}

In order to test our QASS prototype, several functional toy theories for TBR were developed as alternatives to the expensive MC model. By far the most robust of these was the following sinusoidal theory with adjustable wavenumber parameter $n$:

@@

plotted in~\fref{fig:sintoy} for $n=1$ and two continuous parameters $C$.
ANNs trained on this model demonstrated similar performance to those on the expensive
MC model. QASS performance was verified by training a $\text{1h3f}(256)$ ANN on
the sinusoidal theory for varied quantities of initial, incremental, and MCMC
candidate samples. Although the scope of this project did not include thorough
searches of this hyperparameter domain, sufficient runs were made to identify
some likely trends.

An increase in initial samples with increment held constant had a strong impact
on final surrogate precision, an early confirmation of basic functionality. An
increase in MCMC candidate samples was seen to have a positive but very weak
effect on final surrogate precision, suggesting that the runtime of MCMC on each
iteration can be limited for increased efficiency. The most complex dynamics
arose with the adjustment of sample increment, shown in~\fref{fig:qassincr}. For
each tested initial sample quantity $N$, the optimal number of step samples was seen to be well-approximated by $\sqrt{N}$. The plotted error trends suggest that
incremental samples larger than this optimum give slower model improvement on
both the training and evaluation sets, and a larger minimum error on the
evaluation set. This performance distinction is predicted to be even more
significant when trained on the expensive MC model, where the number of sample
evaluations will serve as the primary bottleneck for computation time.

\begin{figure}
	\centering
	\includegraphics[width=\linewidth]{fig6a_qassincrsamp.pdf}
	\includegraphics[width=\linewidth]{fig6b_qassincrtime.pdf}
    \caption{QASS absolute training error over total sample quantity (left) and number of iterations (right). MAE represents surrogate error on the adaptively-sampled training/test set, and E\_MAE on the independent evaluation sets.}
    \label{fig:qassincr}
\end{figure}

The plateau effect in surrogate error on the evaluation set, seen
in~\fref{fig:qassincr}, was universal to all configurations and thought to
warrant further investigation. At first this was suspected to be a residual
effect of retraining the same ANN instance without adjustment to data
normalisation. A ``Goldilocks scheme'' for checking normalisation drift was
implemented and tested, but did not affect QASS performance. Schemes in which
the ANN is periodically retrained were also discarded, as the retention of
network weights from one iteration to the next was demonstrated to greatly
benefit QASS efficiency. Further insight came from direct comparison between
QASS and a baseline scheme with uniformly random incremental samples, shown
in~\fref{fig:qasssampling}.

>>>>

@@

\subsection{Adaptive Approach}\label{sec:adaptiveres}

In order to test our QASS prototype, several functional toy theories for TBR were developed as alternatives to the expensive MC model. By far the most robust of these was the following sinusoidal theory over continuous parameters $C$, with adjustable wavenumber parameter $n$:

@@

ANNs trained on this model demonstrated similar performance to those on the expensive
MC model. QASS performance was verified by training a $\text{1h3f}(256)$ ANN on
the sinusoidal theory for varied quantities of initial, incremental, and MCMC
candidate samples.

An increase in initial samples with increment held constant had a strong impact
on final surrogate precision, an early confirmation of basic functionality. An
increase in MCMC candidate samples was seen to have a positive but very weak
effect on final surrogate precision, suggesting that the runtime of MCMC on each
iteration could be limited for increased efficiency. We also found that an optimum increment exists and is monotonic with initial sample quantity, above or below which models showed slower improvement on both the training and evaluation sets, and a larger minimum error on the
evaluation set. This performance distinction will be far more
significant for an expensive model such as Paramak, where the number of sample
evaluations is the primary computational bottleneck.

A plateau effect in surrogate error on the evaluation set was universal to all configurations, and initially suspected to be a residual
effect of retraining the same ANN instance without adjustment to data
normalisation. A ``Goldilocks scheme'' for checking normalisation drift was
implemented and tested, but did not affect QASS performance. Schemes in which
the ANN is periodically retrained were also discarded, as the retention of
network weights from one iteration to the next was demonstrated to greatly
benefit QASS efficiency. Further insight came from direct comparison between
QASS and a baseline scheme with uniformly random incremental samples, shown
in~\fref{fig:qasssampling}.

### 2020-05-29 paper/ucl_tbr_paper.tex
more edits to the methodology section to comply with journal guidelines
shorten the first two results sections, transition to two-column layout
Reduce introduction and adaptive sampling sections.

\usepackage{siunitx}
\usepackage[colorinlistoftodos,textsize=tiny]{todonotes} % TODO: remove me before submitting
\usepackage{isotope}
\usepackage{fixltx2e}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{url}

@@

@@

\author{G~Van Goffrier$^1$ and P~Mánek$^{1,2}$, V~Gopakumar$^3$, N~Nikolau$^1$, J~Shimwell$^3$, I~Waldmann$^1$}

@@

\eads{\mailto{graham.vangoffrier.19@ucl.ac.uk}, \mailto{petr.manek.19@ucl.ac.uk}}

@@

% TODO: revise these
\keywords{magnetic moment, solar neutrinos, astrophysics}
\submitto{\jpg}
\maketitle

>>>>

\usepackage{siunitx}
\usepackage[colorinlistoftodos,textsize=tiny]{todonotes} % TODO: remove me before submitting
\usepackage{isotope}
\usepackage{fixltx2e}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{url}
\usepackage{graphicx}

@@

\graphicspath{{./fig/}}

@@

\author{P~Mánek$^{1,2}$, G~Van Goffrier$^1$, V~Gopakumar$^3$, N~Nikolau$^1$, J~Shimwell$^3$ and I~Waldmann$^1$}

@@

\eads{\mailto{petr.manek.19@ucl.ac.uk}, \mailto{graham.vangoffrier.19@ucl.ac.uk}}

@@

% TODO: revise these
\keywords{machine learning, surrogate model, nuclear fusion, tritium breeding, adaptive sampling}
\submitto{\NF}
\maketitle
\ioptwocol
